On Optimality of Finite Element Integration

### 2015-11-23 TOMS_Paper/assembly.tex
BIG CHANGE IN PROGRESS: towards opt + quasi-opt

However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? In this paper, we formulate an approach to solve this problem. Summarizing, our contributions are as follows:

\begin{itemize}
\item We formalize optimality for finite element integration and we build the space of legal transformations within which the optimal is sought.
\item We provide an algorithm to select optimal points in the transformation space. The algorithm uses a cost model to: 1) understand whether a transformation reduces or increases the operation count; 2) choose between different (non-composable) transformations.
\item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

\section{Transformation Space}
\label{sec:optimal-impl}

In this section, we characterize optimality for finite element integration and the transformation space that we need to explore to achieve it.

@@

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions). In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$. An expression (right hand side of a statement) implements a set of monomials. The expression shown is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. We do not pose particular restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and matrix operators, so $\sigma_{ei}$ can be arbitrarily complex. We observe that translating a monomial in matrix notation into code naturally results in this normal form; this applies to any problem, given the rigid structure of a monomial (summarized in Section~\ref{sec:monomials}). Henceforth, without loss of generality and for ease of treatise, we assume finite element integration loop nests characterized by expressions in normal form.

@@

Our aim is a strategy for reducing finite element integration loop nests to \textit{optimal form}. In particular, we characterize optimality as follows.

@@

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of such a transformation space. We consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.

@@

In this section, we study optimal \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (if necessary) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we show that by restricting ourselves to the class of loops and expressions constructed in Section~\ref{sec:lnopt}, the search space is actually very small.

@@

\subsubsection{Reduction and free order loops}
\label{sec:se-rln}
The arithmetic complexity of $\sigma_{ei}$ varies with the differential and vector operators employed in a monomial, becoming critical in complicated formulations (characterized, for instance, by several, possibly non-linear coefficients). We distinguish the following cases.
\begin{enumerate}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$. This is the simplest scenario since factorizing $f_i$ has no side effects.
\item There is no common symbol among all terms, although there still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$. Factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost (i.e., code motion opportunities are destroyed).
\item Within a term $\sigma_{ei}^{w}$, sharing elimination could still be performed by (i) factorizing integration-dependent symbols and (ii) hoisting geometry-dependent sub-expressions. This is tempting, but not necessarily the optimal choice depending on two factors: the trip count of $L_i$ and the arithmetic complexity of $\sigma_{ei}^{w}$. We provide an example. Consider the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient}. We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1. We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required. As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics.
\end{enumerate}

@@

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

@@

% matrix notation, fixed normal form, so no need for fancy algors to do CSE
\end{Algo}

@@

\section{Optimal Synthesis}
\label{sec:optimal-synthesis}
In this section, we build a transformation algorithm capable of reaching the goal set by Definition~\ref{def:optimality} for the class of loop nests in Definition~\ref{def:fem-loopnest}. In the perspective of integrating the transformation algorithm with a code generation system, we also discuss an approach that trades optimality for compilation time.

>>>>

However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? In this paper, we formulate an approach that solves this problem for a particular class of forms and provides very good approximations (``quasi-optimality'') in all other cases. Summarizing, our contributions are as follows:

\begin{itemize}
\item We formalize optimality for finite element integration and we build the space of legal transformations within which the optimal is sought.
\item We provide an algorithm to select points in the transformation space. The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
\item We explain under what conditions our algorithm leads to optimality. In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) in general, quasi-optimality is at least achieved (i.e., operation count optimal for innermost loops, close-to-optimal for outer loops).
\item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

\section{Transformation Space}
\label{sec:optimal-impl}
In this section, we characterize optimality for finite element integration and the transformation space that we need to explore to achieve it.

@@

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions). In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$. An expression (right hand side of a statement) implements a set of monomials. The expression shown is in \textit{canonical form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. We do not pose particular restrictions on function spaces (e.g., scalar- or vector-valued), coefficients (e.g., linear or non-linear), differential and vector operators, so $\sigma_{ei}$ can be arbitrarily complex. We observe that translating a monomial in matrix notation into code naturally leads to the canonical form. This applies to any problem, given the rigid structure of a monomial (summarized in Section~\ref{sec:monomials}). Henceforth, without loss of generality and for ease of treatise, we assume finite element integration loop nests with expressions in canonical form.

@@

For this class of loop nests and expressions, we characterize optimality as follows.

@@

Now that we have established the concepts of optimality and finite element integration loop nest, we can elaborate on goals and results of this article. The main objective is a strategy to systematically achieve optimality -- as formalized in Definition~\ref{def:mln-optimality} -- in classes of finite element integration loop nests -- as formalized in Definition~\ref{def:fem-loopnest}. For instance, a class of loop nests may be characterized by the use of only certain differential or tensor operators, or by certain properties in the resulting expressions. What our approach manages to achieve is: (i) optimality whenever the $\sigma_{ei}$  expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases. With quasi-optimality we mean that the operation count is minimal in the innermost loop and close to the ideal one in all outer loops.

Our approach needs:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item an algorithm to select points in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space and on understanding the conditions for optimality. We will show that whenever an expression (i.e., its monomials) is ``too unstructured'', optimality may not be reached (unless, obviously, using brute-force based approaches). We will consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to minimize the operation count by applying factorization, code motion and (if necessary) common sub-expression elimination to remove sharing. In generic problems, transformations of this sort usually require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we show that since we restrict ourselves to a specific class of loops and expressions (described in Section~\ref{sec:lnopt}), we can formulate a simple algorithm ensuring quasi-optimality in general and optimality when expressions have a particular structure (typical of many finite element problems).

@@

\subsubsection{Reduction and free order loops}
\label{sec:se-rln}
The arithmetic complexity of $\sigma_{ei}$ varies with the differential and vector operators employed in a monomial, becoming critical in complicated formulations (characterized, for instance, by several, possibly non-linear coefficients). We distinguish the following cases.
\begin{enumerate}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the canonical form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$. This is the simplest scenario since factorizing $f_i$ has no side effects.
\item There is no common symbol among all terms, although there still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$. Factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost (i.e., code motion opportunities are destroyed).
\item Within a term $\sigma_{ei}^{w}$, sharing elimination could still be performed by (i) factorizing integration-dependent symbols and (ii) hoisting geometry-dependent sub-expressions. This is tempting, but not necessarily the optimal choice depending on two factors: the trip count of $L_i$ and the arithmetic complexity of $\sigma_{ei}^{w}$. We provide an example. Consider the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient}. We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1. We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required. As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics.
\end{enumerate}

@@

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in the canonical form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

@@

% matrix notation, fixed canonical form, so no need for fancy algors to do CSE
\end{Algo}

@@

\section{Optimal Synthesis}
\label{sec:optimal-synthesis}
In this section, we build a transformation algorithm capable of reaching optimality (Definition~\ref{def:optimality}) in a class of finite element integration loop nests (Definition~\ref{def:fem-loopnest}) and quasi-optimality in any other finite element integration loop nest. In the perspective of integrating the transformation algorithm with a code generation system, we also discuss an approach that trades optimality for compilation time.
